---
title: "version 1"
output:
  html_document: default
  word_document: default
date: "2024-10-07"
---

-

<<<<<<< HEAD

=======
HIV_limited data dictionary:

newid - subject id 
RACE - subject race
EDUCBAS - subject education at baseline 
drug_use_status - subject drug group
drug_use_currently - current drug use, 1 = yes, 0 = no
drug_use_previously - previous drug use, 1 = yes, 0 = no
drug_use_never - never drug use, 1 = yes, 0 = no
AGG_MENT_0 - SF36 MCS score year 0
AGG_MENT_2 - SF36 MCS score year 2
AGG_PHYS_0 - SF36 PCS score year 0
AGG_PHYS_2 - SF36 PCS score year 2
HASHV_2 - Hash use since last visit year 2
HASHF_2 - Frequency of hash use since last visit year 2
income_2 - income year 2
BMI_2 - bmi year 2
HBP_2 - high blood pressure year 2
DIAB_2 - diabetes year 2
LIV34_2 - liver disease stage 3 or 4 year 2
KID_2 - kidney disease year 2
FRP_2 - Frailty Related Phenotype year 2
FP_2 - Frailty Phenotype year 2
TCHOL_2 - total cholesterol year 2
TRIG_2 - triglycerides year 2
LDL_2 - Low Density Lipoprotein year 2
DYSLIP_2 - dyslipidemia at visit year 2
CESD_2 - Depression scale year 2
SMOKE_2 - smoking status year 2
DKGRP_2 - Alcohol use since last visit year 2
HEROPIATE_2 - Heroin use year 2
IDU_2 - Took/used drugs with a needle since last visit? year 2
LEU3N_0 - # of CD4 positive cells year 0
LEU3N_2 - # of CD4 positive cells year 2
VLOAD_0 - viral load year 0
VLOAD_2 - viral load year 2
ADH_2 - adherence year 2
age_2 - age year 2
ART_2 - take art at year 2 visit
RACE_2 - race, white Hispanic 1 = yes 0 = no
RACE_3 - race, black non-Hispanic 1 = yes 0 = no
RACE_4 - race, black Hispanic 1 = yes 0 = no
RACE_7 - race, other 1 = yes 0 = no
RACE_8 - race, other Hispanic 1 = yes 0 = no
EDUCBAS_2 - education at baseline, 9, 10, or 11th grade, 1 = yes 0 = no
EDUCBAS_3 - education at baseline, 12th grade, 1 = yes 0 = no
EDUCBAS_4 - education at baseline, At least one year college but no degree, 1 = yes 0 = no
EDUCBAS_5 - education at baseline, Four years college / got degree, 1 = yes 0 = no
EDUCBAS_6 - education at baseline, Some graduate work, 1 = yes 0 = no
EDUCBAS_7 - education at baseline, Post-graduate degree, 1 = yes 0 = no
HASHF_2_1 - Frequency of hash use year 2, daily, 1 = yes 0 = no
HASHF_2_2 - Frequency of hash use year 2, weekly, 1 = yes 0 = no
HASHF_2_3 - Frequency of hash use year 2, monthly, 1 = yes 0 = no
HASHF_2_4 - Frequency of hash use year 2, less often, 1= yes 0 = no
income_2_2 - income year 2, 10,000-19,999, 1 = yes 0 = no
income_2_3 - income year 2, 20,000-29,999, 1 = yes 0 = no
income_2_4 - income year 2, 30,000-39,999, 1 = yes 0 = no
income_2_5 - income year 2, 40,000-49,999, 1 = yes 0 = no
income_2_6 - income year 2, 50,000-59,999, 1 = yes 0 = no
income_2_7 - income year 2, 60,000 or more, 1 = yes 0 = no
income_2_9 - income year 2, do not wish to answwer, 1 = yes 0 = no
HBP_2_2 - High blood pressure year 2, yes, 1 = yes 0 = no 
HBP_2_3 - High blood pressure year 2, No, based data trajectory, 1 = yes 0 = no
HBP_2_4 - High blood pressure year 2, yes, based data trajectory, 1 = yes 0 = no
HBP_2_9 - High blood pressure year 2, insufficient data, 1 = yes 0 = no
DIAB_2_2 - diabetes year 2, yes, 1 = yes 0 = no 
DIAB_2_3 - diabetes year 2, No, based data trajectory, 1 = yes 0 = no
DIAB_2_9 - diabetes year 2, insufficient data, 1 = yes 0 = no
LIV34_2_2 - stage 3 or 4 liver disease year 2, yes, 1 = yes 0 = no 
LIV34_2_9 - stage 3 or 4 liver disease year 2, insufficient data, 1 = yes 0 = no
KID_2_2 - kidney disease at year 2, yes, 1 = yes, 0 = no
KID_2_9 - kidney disease at year 2, insufficient data, 1 = yes 0 = no
FRP_2_2 - Frailty Related Phenotype year 2, yes, 1 = yes 0 = no
FRP_2_9 - Frailty Related Phenotype year 2, insufficient data, 1 = yes 0 = no
FP_2_2 - Frailty Phenotype year 2, yes, 1 - yes 0 = no
FP_2_9 - Frailty Phenotype year 2, insufficient data, 1 - yes 0 = no
DYSLIP_2_2 - dyslipidemia at visit, yes, 1 = yes 0 = no
DYSLIP_2_3 - dyslipidemia at visit, No, based data trajectory, 1 = yes 0 = no
DYSLIP_2_4 - dyslipidemia at visit, yes, based data trajectory, 1 = yes 0 = no
DYSLIP_2_9 - dyslipidemia at visit, insufficient data, 1 = yes 0 = no
SMOKE_2_2 - smoking status year 2, former smoker, 1 = yes, 0 = no
SMOKE_2_3 - smoking status year 2, current smoker, 1 = yes, 0 = no
DKGRP_2_1 - alcohol since last visit year 2, 1-3 drinks per week, 1 = yes, 0 = no
DKGRP_2_2 - alcohol since last visit year 2, 4-13 drinks per week, 1 = yes, 0 = no
DKGRP_2_3 - alcohol since last visit year 2, 13 or more drinks per week, 1 = yes, 0 = no
ADH_2_2 - adherence to treatment year 2, 95-99%, 1 = yes 0= no
ADH_2_3 - adherence to treatment year 2, 75-94%, 1 = yes 0= no
ADH_2_4 - adherence to treatment year 2, under 75%, 1 = yes 0= no



#some steps for commiting to git and pushing to github:
>>>>>>> 311c46439c0dfe67b62173d3607a3ac04778134f
for git use getwd() function to determine pathway 
cd /home/robmcneil/Desktop/test
git branch
git checkout your-branch-name (can do this in R terminal )
git status
git push origin your-branch-name




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, include= FALSE}

#read in data

library(tidyverse)
library(dplyr)
library(ggplot2)
library(VIM)
library(naniar)
library(readr)
library(kableExtra)
library(finalfit)
library(broom)
library(sjPlot)
library(StepReg)
library(reshape2)
library(mice)
library(fastDummies)
library(corrplot)


#linux pathway
#hiv<- read_csv("/home/robmcneil/Documents/advanced_data/hiv_dataset.csv")

#windows pathway
hiv<- read_csv("C:/Users/rrm10/OneDrive/Desktop/advanced data analysis/hiv_dataset.csv")

# Jessicas read in data
#hiv<- read_csv("~/Downloads/hiv_dataset.csv")

#check data types with str function
str(hiv)

```



```{r, include=FALSE}
#filter dataset to two years
print(hiv)
hiv_filtered <- hiv %>%
  filter(years <= 2)
print(hiv_filtered)
```

```{r, include=FALSE}
# Get drug use status
hiv_categorized <- hiv_filtered %>%
  group_by(newid) %>%
  mutate(drug_use_status = case_when(
    all(hard_drugs == 0) ~ "Never",
    hard_drugs[3] == 1 ~ "Current",
    hard_drugs[3] == 0 & any(hard_drugs[1:2] == 1) ~ "Previous"
  )) %>%
  
  ungroup()
  
print(hiv_categorized)

```

```{r}
#run initial correlation matrix to see how outcome variables are correlated with eachother

# Run the correlation on numeric columns only
correlation_matrix <- cor(hiv_filtered %>% select(where(is.numeric)), use = "complete.obs")



# Print compact correlation matrix
kable(correlation_matrix, digits = 2, caption = "Correlation Matrix for HIV Data (Years <= 2)") %>%
  kable_styling(position = "center", full_width = F)


```


```{r, include=FALSE}
#Get dummy variables for drug use status 

hiv_categorized <- hiv_categorized %>%
  mutate(
    drug_use_currently = if_else(drug_use_status == "Current", 1, 0),
    drug_use_previously = if_else(drug_use_status == "Previous", 1, 0),
    drug_use_never = if_else(drug_use_status == "Never", 1, 0)
  )
print(hiv_categorized)

```


```{r, include= FALSE}
#Turn dataset from long to wide for missing data evaluation, so we only have one observation per subject

wide_hiv_categorized <-hiv_categorized %>%
  pivot_wider(
    id_cols = c(newid, RACE, EDUCBAS, hivpos, drug_use_status, drug_use_currently, drug_use_previously, drug_use_never),
    names_from = years,
    values_from = c(AGG_MENT, AGG_PHYS, HASHV, HASHF, income, BMI, HBP, DIAB,LIV34,KID,FRP,FP, TCHOL,TRIG,
                    LDL,DYSLIP,CESD,SMOKE,DKGRP,HEROPIATE,IDU,LEU3N,VLOAD,ADH,age, ART, everART, 
                    hard_drugs)
  )

print(wide_hiv_categorized)

```

```{r,echo=FALSE}
# Visualize what variables are missing data and how much is missing  
aggr(wide_hiv_categorized, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(wide_hiv_categorized), cex.axis=0.7, gap=3, ylab=c("Missing data","Pattern"))

vis_miss(wide_hiv_categorized)
```


```{r,echo=FALSE}
#See what percent of data is missing for non-outcome variables by drug use status
wide_hiv_categorized$TRIG_1_missing <- ifelse(is.na(wide_hiv_categorized$TRIG_1), 1, 0)
wide_hiv_categorized$LDL_1_missing <- ifelse(is.na(wide_hiv_categorized$LDL_1),1,0)
wide_hiv_categorized$TRIG_0_missing <- ifelse(is.na(wide_hiv_categorized$TRIG_0), 1, 0)
wide_hiv_categorized$LDL_2_missing <- ifelse(is.na(wide_hiv_categorized$LDL_2), 1, 0)
wide_hiv_categorized$LDL_0_missing <- ifelse(is.na(wide_hiv_categorized$LDL_0), 1, 0)
wide_hiv_categorized$TRIG_2_missing <- ifelse(is.na(wide_hiv_categorized$TRIG_2), 1, 0)
wide_hiv_categorized$TCHOL_1_missing <- ifelse(is.na(wide_hiv_categorized$TCHOL_1), 1, 0)
wide_hiv_categorized$TCHOL_0_missing <- ifelse(is.na(wide_hiv_categorized$TCHOL_0), 1, 0)
wide_hiv_categorized$TCHOL_2_missing <- ifelse(is.na(wide_hiv_categorized$TCHOL_2), 1, 0)
wide_hiv_categorized$income_2_missing <- ifelse(is.na(wide_hiv_categorized$income_2), 1, 0)
wide_hiv_categorized$income_0_missing <- ifelse(is.na(wide_hiv_categorized$income_0), 1, 0)
wide_hiv_categorized$income_1_missing <- ifelse(is.na(wide_hiv_categorized$income_1), 1, 0)
wide_hiv_categorized$BMI_2_missing <- ifelse(is.na(wide_hiv_categorized$BMI_2), 1, 0)
wide_hiv_categorized$BMI_1_missing <- ifelse(is.na(wide_hiv_categorized$BMI_1), 1, 0)
wide_hiv_categorized$HASHF_0_missing <- ifelse(is.na(wide_hiv_categorized$HASHF_0), 1, 0)
wide_hiv_categorized$VLOAD_1_missing <- ifelse(is.na(wide_hiv_categorized$VLOAD_1), 1, 0)
wide_hiv_categorized$LEU3N_1_missing <- ifelse(is.na(wide_hiv_categorized$LEU3N_1), 1, 0)
wide_hiv_categorized$HASHV_2_missing <- ifelse(is.na(wide_hiv_categorized$HASHV_2), 1, 0)
wide_hiv_categorized$HASHV_1_missing <- ifelse(is.na(wide_hiv_categorized$HASHV_1), 1, 0)
wide_hiv_categorized$BMI_0_missing <- ifelse(is.na(wide_hiv_categorized$BMI_0), 1, 0)
wide_hiv_categorized$HASHF_1_missing <- ifelse(is.na(wide_hiv_categorized$HASHF_1), 1, 0)
wide_hiv_categorized$DKGRP_1_missing <- ifelse(is.na(wide_hiv_categorized$DKGRP_1), 1, 0)
wide_hiv_categorized$DKGRP_2_missing <- ifelse(is.na(wide_hiv_categorized$DKGRP_2), 1, 0)
wide_hiv_categorized$HEROPIATE_0_missing <- ifelse(is.na(wide_hiv_categorized$HEROPIATE_0), 1, 0)
wide_hiv_categorized$HEROPIATE_1_missing <- ifelse(is.na(wide_hiv_categorized$HEROPIATE_1), 1, 0)
wide_hiv_categorized$AGG_MENT_1_missing <- ifelse(is.na(wide_hiv_categorized$AGG_MENT_1), 1, 0)
wide_hiv_categorized$AGG_PHYS_1_missing <- ifelse(is.na(wide_hiv_categorized$AGG_PHYS_1), 1, 0)


missing_demogrphic_by_drug_use_status <- wide_hiv_categorized %>%
  group_by(drug_use_status) %>%
  summarize(
    percent_missing_TRIG_1 = mean(TRIG_1_missing) * 100,
    percent_missing_LDL_1 = mean (LDL_1_missing) * 100,
    percent_missing_TRIG_0 = mean (TRIG_0_missing) *100,
    percent_missing_LDL_2 = mean (LDL_2_missing) *100,
    percent_missing_LDL_0 = mean (LDL_0_missing) *100,
    percent_missing_TRIG_2 = mean (TRIG_2_missing) *100,
    percent_missing_TCHOL_1 = mean (TCHOL_1_missing) *100,
    percent_missing_TCHOL_0 = mean(TCHOL_0_missing) * 100,
    percent_missing_TCHOL_2 = mean (TCHOL_2_missing) * 100,
    percent_missing_income_2 = mean (income_2_missing) *100,
    percent_missing_income_0 = mean (income_0_missing) *100,
    percent_missing_income_1 = mean (income_1_missing) *100,
    percent_missing_BMI_2 = mean (BMI_2_missing) *100,
    percent_missing_BMI_1 = mean (BMI_1_missing) *100,
    percent_missing_HASHF_0 = mean (HASHF_0_missing) *100,
    percent_missing_VLOAD_1 = mean(VLOAD_1_missing) * 100,
    percent_missing_LEU3N_1 = mean (LEU3N_1_missing) * 100,
    percent_missing_HASHV_2 = mean (HASHV_2_missing) *100,
    percent_missing_HASHV_1 = mean (HASHV_1_missing) *100,
    percent_missing_BMI_0 = mean (BMI_0_missing) *100,
    percent_missing_HASHF_1 = mean (HASHF_1_missing) *100,
    percent_missing_DKGRP_1 = mean (DKGRP_1_missing) *100,
    percent_missing_DKGRP_2 = mean (DKGRP_2_missing) *100,
    percent_missing_HEROPIATE_0 = mean (HEROPIATE_0_missing) *100,
    percent_missing_HEROPIATE_1 = mean (HEROPIATE_1_missing) *100,
    percent_missing_AGG_MENT_1 = mean (AGG_MENT_1_missing) *100,
    percent_missing_AGG_PHYS_1 = mean (AGG_PHYS_1_missing) *100,
  )

print(missing_demogrphic_by_drug_use_status)
```

```{r,echo=FALSE}
#see what percent of of each outcome variable is missing by drug use status group
wide_hiv_categorized$VLOAD_0_missing <- ifelse(is.na(wide_hiv_categorized$VLOAD_0), 1, 0)
wide_hiv_categorized$VLOAD_2_missing <- ifelse(is.na(wide_hiv_categorized$VLOAD_2),1,0)
wide_hiv_categorized$LEU3N_0_missing <- ifelse(is.na(wide_hiv_categorized$LEU3N_0), 1, 0)
wide_hiv_categorized$LEU3N_2_missing <- ifelse(is.na(wide_hiv_categorized$LEU3N_2), 1, 0)
wide_hiv_categorized$AGG_MENT_0_missing <- ifelse(is.na(wide_hiv_categorized$AGG_MENT_0), 1, 0)
wide_hiv_categorized$AGG_MENT_2_missing <- ifelse(is.na(wide_hiv_categorized$AGG_MENT_2), 1, 0)
wide_hiv_categorized$AGG_PHYS_0_missing <- ifelse(is.na(wide_hiv_categorized$AGG_PHYS_0), 1, 0)
wide_hiv_categorized$AGG_PHYS_2_missing <- ifelse(is.na(wide_hiv_categorized$AGG_PHYS_2), 1, 0)
wide_hiv_categorized$ADH_2_missing <- ifelse(is.na(wide_hiv_categorized$ADH_2), 1, 0)
  
missing_percent_by_drug_use_status <- wide_hiv_categorized %>%
  group_by(drug_use_status) %>%
  summarize(
    percent_missing_VLOAD_0 = mean(VLOAD_0_missing) * 100,
    percent_missing_VLOAD_2 = mean (VLOAD_2_missing) * 100,
    percent_missing_LEU3N_0 = mean (LEU3N_0_missing) *100,
    percent_missing_LEU3N_2 = mean (LEU3N_2_missing) *100,
    percent_missing_AGG_MENT_0 = mean (AGG_MENT_0_missing) *100,
    percent_missing_AGG_MENT_2 = mean (AGG_MENT_2_missing) *100,
    percent_missing_ADH_2 = mean (ADH_2_missing) *100,
    percent_missing_AGG_PHYS_0 = mean (AGG_PHYS_0_missing) *100,
    percent_missing_AGG_PHYS_2 = mean (AGG_PHYS_2_missing) *100,
  )

print(missing_percent_by_drug_use_status)
```
visualize percent missing by drug use status

```{r}
# Reshape data for ggplot
missing_long <- missing_percent_by_drug_use_status %>%
  pivot_longer(cols = starts_with("percent_missing"), 
               names_to = "outcome", 
               values_to = "percent_missing")

# Create the bar plot
ggplot(missing_long, aes(x = outcome, y = percent_missing, fill = drug_use_status)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Figure 5:Percentage of Missing Data by Outcome and Drug Use Status",
       x = "Outcome Variables",
       y = "Percentage Missing") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))





```
Important finding: that outcome variables are only missing for those in the never treatment group, data is missing at random 


```{r,echo=FALSE}
#visualize data that is missing jst in the never group to look for further patterns
data_never <- wide_hiv_categorized %>%
  filter(drug_use_status == "Never")

aggr(data_never, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(data_never), cex.axis=0.7, gap=3, ylab=c("Missing data","Pattern"))

vis_miss(data_never)
```

```{r,echo=FALSE}
#visualize similarities between missingness
missing_data <- wide_hiv_categorized %>%
  filter(is.na(VLOAD_2))

print(missing_data)

gg_miss_upset(missing_data)
```


```{r,include=FALSE}


# Filter the dataset for observations where drug_use_status == "Never"

data_previous <- wide_hiv_categorized %>%
  filter(drug_use_status == "Previous")
data_current <- wide_hiv_categorized %>%
  filter(drug_use_status == "Current")

data_never$log_vload2 <- log(data_never$VLOAD_2 +1)
data_never$log_vload0 <- log(data_never$VLOAD_0 +1)
data_previous$log_vload0 <- log(data_previous$VLOAD_0 +1)
data_previous$log_vload2 <- log(data_previous$VLOAD_2 +1)
data_current$log_vload0 <- log(data_current$VLOAD_0 +1)
data_current$log_vload2 <- log(data_current$VLOAD_2 +1)
# Loop through the numeric columns and create histograms
numeric_vars <- names(data_never)[sapply(data_never, is.numeric)]

vars <- c("VLOAD_0","VLOAD_1","VLOAD_2" )
for (var in vars) {
  # Create the histogram
  p <- ggplot(data_never, aes_string(x = var), breaks=10) +
    geom_histogram(breaks = 10, fill = "skyblue", color = "black") +
    labs(title = paste("Histogram of", var, "for Never Drug Use Status")) +
    theme_minimal()
  
  # Print each plot
  print(p)

}
hist(data_never$log_vload0, breaks = 20, main = "Histogram of Numeric Variable", xlab = "Values", ylab = "Frequency")

hist(data_never$log_vload2, breaks = 20, main = "Histogram of Numeric Variable", xlab = "Values", ylab = "Frequency")

hist(data_previous$log_vload0, breaks = 20, main = "Histogram of Numeric Variable", xlab = "Values", ylab = "Frequency")

hist(data_previous$log_vload2, breaks = 20, main = "Histogram of Numeric Variable", xlab = "Values", ylab = "Frequency")

hist(data_current$log_vload0, breaks = 20, main = "Histogram of Numeric Variable", xlab = "Values", ylab = "Frequency")

hist(data_current$log_vload2, breaks = 20, main = "Histogram of Numeric Variable", xlab = "Values", ylab = "Frequency")
```
 



```{r, include=FALSE}
#impute data using knn try 2 (this one is better)
HIV_imputed1 <- kNN(wide_hiv_categorized)
print(HIV_imputed1)
```
```{r, include=FALSE}
#choose certain variables so dataset is less overwhelming 
HIV_limited <- HIV_imputed1 %>%
  select(newid,RACE,EDUCBAS,drug_use_status,drug_use_currently,drug_use_previously,drug_use_never, AGG_MENT_0,AGG_MENT_2,AGG_PHYS_0, AGG_PHYS_2, HASHV_2, HASHF_2, income_2,  BMI_2, HBP_2, DIAB_2, LIV34_2, KID_2, FRP_2,FP_2, TCHOL_2,TRIG_2,LDL_2, DYSLIP_2, CESD_2, SMOKE_2, DKGRP_2, HEROPIATE_2, IDU_2, LEU3N_0,LEU3N_2, VLOAD_0, VLOAD_2, ADH_0, ADH_2, age_2, ART_2, everART_0)
print(HIV_limited)
```
```{r, include=FALSE}
#Convert numeric categorical variables to character so you can get dummies 
HIV_limited[, c("RACE","EDUCBAS","HASHF_2", "income_2","HBP_2","DIAB_2","LIV34_2","KID_2", "FRP_2","FP_2","DYSLIP_2","SMOKE_2","DKGRP_2","ADH_0","ADH_2")] <- lapply(HIV_limited[,c("RACE","EDUCBAS","HASHF_2", "income_2","HBP_2","DIAB_2","LIV34_2","KID_2", "FRP_2","FP_2","DYSLIP_2","SMOKE_2","DKGRP_2","ADH_0","ADH_2")],as.character)
str(HIV_limited)
```

```{r, include=FALSE}
#Get dummy variables
category_vars <- c("RACE","EDUCBAS","HASHF_2", "income_2","HBP_2","DIAB_2","LIV34_2","KID_2", "FRP_2","FP_2","DYSLIP_2","SMOKE_2","DKGRP_2","ADH_0","ADH_2")
for (var in category_vars) {
  HIV_limited <- dummy_cols(HIV_limited, select_columns = var, remove_first_dummy = TRUE)
}
print(HIV_limited)
```

```{r,include=FALSE}

#code for table 1 
demographic_summary <-  HIV_limited %>% 
  group_by(drug_use_status) %>%
  summarise(
    `White, non-Hispanic (%)` = as.character(round(sum(RACE == 1, na.rm = TRUE) / n() * 100, 2)),
    `White, Hispanic (%)` = as.character(round(sum(RACE == 2, na.rm = TRUE) / n() * 100, 2)),
    `Black, non-Hispanic (%)` = as.character(round(sum(RACE == 3, na.rm = TRUE) / n() * 100, 2)),
    `Black, Hispanic (%)` = as.character(round(sum(RACE == 4, na.rm = TRUE) / n() * 100, 2)),
    `American Indian or Alaskan Native (%)` = as.character(round(sum(RACE == 5, na.rm = TRUE) / n() * 100, 2)),
    `Asian or Pacific Islander (%)` = as.character(round(sum(RACE == 6, na.rm = TRUE) / n() * 100, 2)),
    `Other (%)` = as.character(round(sum(RACE == 7, na.rm = TRUE) / n() * 100, 2)),
    `Other Hispanic (%)` = as.character(round(sum(RACE == 8, na.rm = TRUE) / n() * 100, 2)),
    `Education Level at Baseline: 8th grade or less (%)` = as.character(round(sum(EDUCBAS == 1, na.rm = TRUE) / n() * 100, 2)),
    `Education Level at Baseline: 9, 10, or 11th grade (%)` = as.character(round(sum(EDUCBAS == 2, na.rm = TRUE) / n() * 100, 2)),
    `Education Level at Baseline: 12th grade (%)` = as.character(round(sum(EDUCBAS == 3, na.rm = TRUE) / n() * 100, 2)),
    `Education Level at Baseline: At least 1 year of college - no degree (%)` = as.character(round(sum(EDUCBAS == 4, na.rm = TRUE) / n() * 100, 2)),
    `Education Level at Baseline: 4 years college / got degree (%)` = as.character(round(sum(EDUCBAS == 5, na.rm = TRUE) / n() * 100, 2)),
    `Education Level at Baseline: Some graduate work (%)` = as.character(round(sum(EDUCBAS == 6, na.rm = TRUE) / n() * 100, 2)),
    `Education Level at Baseline: post graduate degree (%)` = as.character(round(sum(EDUCBAS == 7, na.rm = TRUE) / n() * 100, 2)),
    `Individual gross income (year2): Less than $10,000 (%)` = as.character(round(sum(income_2 == 1, na.rm = TRUE) / n() * 100, 2)),
    `Individual gross income (year2): $10,000-$19,999 (%)` = as.character(round(sum(income_2 == 2, na.rm = TRUE) / n() * 100, 2)),
    `Individual gross income (year2): $20,000-29,999 (%)` = as.character(round(sum(income_2 == 3, na.rm = TRUE) / n() * 100, 2)),
    `Individual gross income (year2): $30,000-$39,999 (%)` = as.character(round(sum(income_2 == 4, na.rm = TRUE) / n() * 100, 2)),
    `Individual gross income (year2): $40,000-$49,999 (%)` = as.character(round(sum(income_2 == 5, na.rm = TRUE) / n() * 100, 2)),
    `Individual gross income (year2): $50,000-$59,999 (%)` = as.character(round(sum(income_2 == 6, na.rm = TRUE) / n() * 100, 2)),
    `Individual gross income (year2): $60,000 or more (%)` = as.character(round(sum(income_2 == 7, na.rm = TRUE) / n() * 100, 2)),
    `High Blood Pressure (year2): No (%)` = as.character(round(sum(HBP_2 == 1, na.rm = TRUE) / n() * 100, 2)),
    `High Blood Pressure (year2): Yes (%)` = as.character(round(sum(HBP_2 == 2, na.rm = TRUE) / n() * 100, 2)),
    `High Blood Pressure (year2): No, from data trajectory  (%)` = as.character(round(sum(HBP_2 == 3, na.rm = TRUE) / n() * 100, 2)),
    `High Blood Pressure (year2): Yes, from data trajectory  (%)` = as.character(round(sum(HBP_2 == 4, na.rm = TRUE) / n() * 100, 2)),
    `Diabetes (year2): No (%)` = as.character(round(sum(DIAB_2 == 1, na.rm = TRUE) / n() * 100, 2)),
    `Diabetes (year2): Yes (%)` = as.character(round(sum(DIAB_2 == 2, na.rm = TRUE) / n() * 100, 2)),
    `Diabetes (year2): No, from data trajectory  (%)` = as.character(round(sum(DIAB_2 == 3, na.rm = TRUE) / n() * 100, 2)),
    `Diabetes (year2): Yes, from data trajectory  (%)` = as.character(round(sum(DIAB_2 == 4, na.rm = TRUE) / n() * 100, 2)),
    `Liver Disease Stage 3/4 (year 2): No (%)` = as.character(round(sum(LIV34_2 == 1, na.rm = TRUE) / n() * 100, 2)),
    `Liver Disease Stage 3/4 (year 2) Yes (%)` = as.character(round(sum(LIV34_2 == 2, na.rm = TRUE) / n() * 100, 2)),
    `Kidney Disease (year 2) No (%)` = as.character(round(sum(KID_2 == 1, na.rm = TRUE) / n() * 100, 2)),
    `Kidney Disease (year 2) Yes (%)` = as.character(round(sum(KID_2 == 2, na.rm = TRUE) / n() * 100, 2)),
    `Smoking Status (year 2) Never Smoker (%)` = as.character(round(sum(SMOKE_2 == 1, na.rm = TRUE) / n() * 100, 2)),
    `Smoking Status (year 2) Former Smoker (%)` = as.character(round(sum(SMOKE_2 == 2, na.rm = TRUE) / n() * 100, 2)),
    `Smoking Status (year 2) Current Smoker (%)` = as.character(round(sum(SMOKE_2 == 3, na.rm = TRUE) / n() * 100, 2)),
    `Aggregated Mental Health - Baseline (Mean(SD))` = as.character(paste0(round(mean(AGG_MENT_0, na.rm = TRUE), 2), " (", round(sd(AGG_MENT_0, na.rm = TRUE), 2), ")")),
    `Aggregated Mental Health - Year 2 (Mean(SD))` = as.character(paste0(round(mean(AGG_MENT_2, na.rm = TRUE), 2), " (", round(sd(AGG_MENT_2, na.rm = TRUE), 2), ")")),
    `Aggregated Physical Health - Baseline (Mean(SD))` = as.character(paste0(round(mean(AGG_PHYS_0, na.rm = TRUE), 2), " (", round(sd(AGG_PHYS_0, na.rm = TRUE), 2), ")")),
    `Aggregated Physical Health - Year 2 (Mean(SD))` = as.character(paste0(round(mean(AGG_PHYS_2, na.rm = TRUE), 2), " (", round(sd(AGG_PHYS_2, na.rm = TRUE), 2), ")")),
    `BMI - Year 2 (Mean(SD))` = as.character(paste0(round(mean(BMI_2, na.rm = TRUE), 2), " (", round(sd(BMI_2, na.rm = TRUE), 2), ")")),   
    `Total Cholesterol - Year 2 (Mean(SD))` = as.character(paste0(round(mean(TCHOL_2, na.rm = TRUE), 2), " (", round(sd(TCHOL_2, na.rm = TRUE), 2), ")")), 
    `Triglycerides - Year 2 (Mean(SD))` = as.character(paste0(round(mean(TRIG_2, na.rm = TRUE), 2), " (", round(sd(TRIG_2, na.rm = TRUE), 2), ")")),
    `Depression Scale - Year 2 (Mean(SD))` = as.character(paste0(round(mean(CESD_2, na.rm = TRUE), 2), " (", round(sd(CESD_2, na.rm = TRUE), 2), ")")),
    `Number of CD4 Positive Cells - Year 0 (Mean(SD))` = as.character(paste0(round(mean(LEU3N_0, na.rm = TRUE), 2), " (", round(sd(LEU3N_0, na.rm = TRUE), 2), ")")),
    `Number of CD4 Positive Cells - Year 2 (Mean(SD))` = as.character(paste0(round(mean(LEU3N_2, na.rm = TRUE), 2), " (", round(sd(LEU3N_2, na.rm = TRUE), 2), ")")),
    `Standardized Viral Load - Year 0 (Mean(SD))` = as.character(paste0(round(mean(VLOAD_0, na.rm = TRUE), 2), " (", round(sd(VLOAD_0, na.rm = TRUE), 2), ")")),
    `Standardized Viral Load - Year 2 (Mean(SD))` = as.character(paste0(round(mean(VLOAD_2, na.rm = TRUE), 2), " (", round(sd(VLOAD_2, na.rm = TRUE), 2), ")")),
    `Adherence to Treatment - Year 2 - 100% (%)` = as.character(round(sum(ADH_2 == 1, na.rm = TRUE) / n() * 100, 2)),
    `Adherence to Treatment - Year 2 - 95-99% (%)` = as.character(round(sum(ADH_2 == 2, na.rm = TRUE) / n() * 100, 2)),
    `Adherence to Treatment - Year 2 - 75-94% (%)` = as.character(round(sum(ADH_2 == 3, na.rm = TRUE) / n() * 100, 2)),
    `Adherence to Treatment - Year 2 - <75% (%)` = as.character(round(sum(ADH_2 == 4, na.rm = TRUE) / n() * 100, 2)),
    `Age - Year 2 (Mean(SD))` = as.character(paste0(round(mean(age_2, na.rm = TRUE), 2), " (", round(sd(age_2, na.rm = TRUE), 2), ")")),
  ) %>%
  ungroup() %>%
  mutate(drug_use_status = case_when(
    drug_use_status == "Never" ~ "Never Reports Hard Drug Use",
    drug_use_status == "Previous" ~ "Previously Reports Hard Drug Use",
    drug_use_status == "Current" ~ "Currently Reports Hard Drug Use",
    ))%>%
  pivot_longer(cols = -drug_use_status, names_to = "Demographic and Health", values_to = "Value") %>%
  pivot_wider(names_from = drug_use_status, values_from = Value)

# Print the summary table
kable(demographic_summary, caption = "Table 1: Demographic and Other Characteristic Information by Drug Group")

```
```{r,echo=FALSE}
# Print the summary table 1 with lines between rows
kable(demographic_summary, caption = "Table 1: Demographic and Health Information by Drug Use Status") %>%
  kable_styling() %>%
  row_spec(0, bold = TRUE) %>% # Bold the header
  row_spec(1:nrow(demographic_summary), extra_css = "border-top: 1px solid black; border-bottom: 1px solid black;")
```




```{r,include=FALSE}

#calculate change variables for physical, mental, VLOAD, and LEU3N



hiv_ltd_delta <- HIV_limited %>%
  mutate(
    LEU3N_delta = LEU3N_2 - LEU3N_0,
    VLOAD_delta = VLOAD_0 - VLOAD_2,
    AGG_MENT_delta = AGG_MENT_2 - AGG_MENT_0,
    AGG_PHYS_delta = AGG_PHYS_2 - AGG_PHYS_0
  )







```






```{r,echo=FALSE}
#visualize CD4 positive cells

#test assumption of linearity 
#dont need to test for linearity with categorical variables 
#good graph for presentation 
ggplot(hiv_ltd_delta, aes(x = factor(drug_use_status), y = LEU3N_delta, color = factor(drug_use_status))) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 3, size = 3, color = "black") +  # Adds mean points
  labs(
    title = "Figure 1: Average Change CD4+T Cell based on Drug Usage Groups",
    x = "Drug Usage Group",
    y = "Average Change CD4+ T Cell",
    color = "Treatment Group"
  ) +
  theme_minimal()
```


```{r,echo=FALSE}

#for viral load outcome
#dont need to test for linearity with categorical variables 
#good graph for presentation 
ggplot(hiv_ltd_delta, aes(x = factor(drug_use_status), y = log(VLOAD_delta), color = factor(drug_use_status))) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 3, size = 3, color = "black") +  # Adds mean points
  labs(
    title = "Figure 2: log-transformed Average Change Viral Load based on Drug Usage Groups",
    x = "Drug Usage Group",
    y = "Log Viral Load Average Change",
    color = "Treatment Group"
  ) +
  theme_minimal()




```




```{r,echo=FALSE}
#visualize agg_ment


# Create the box plot
ggplot(hiv_ltd_delta, aes(x = factor(drug_use_status), y = AGG_MENT_delta, color = factor(drug_use_status))) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 3, size = 3, color = "black") +  # Adds mean points
  labs(
    title = "Figure 3: Average Change Mental Health Based on Drug Usage Groups",
    x = "Drug Usage Group",
    y = "AGG_MENT",
    color = "Drug Usage Group"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Current" = "#1f78b4",   # Blue
                                  "Never" = "#33a02c",    # Green
                                  "Previous" = "#e31a1c"))  # Red






```
```{r,echo=FALSE}
#visualize agg_phys
# Create the box plot
ggplot(hiv_ltd_delta, aes(x = factor(drug_use_status), y = AGG_PHYS_delta, color = factor(drug_use_status))) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 3, size = 3, color = "black") +  # Adds mean points
  labs(
    title = "Figure 4: Average Change of Physical Health Based on Drug Usage Groups",
    x = "Drug Usage Group",
    y = "AGG_Phys",
    color = "Drug Usage Group"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Current" = "#1f78b4",   # Blue
                                  "Never" = "#33a02c",    # Green
                                  "Previous" = "#e31a1c"))  # Red





```




```{r,echo=FALSE}

# Select only numeric columns
numeric_columns <- sapply(hiv_ltd_delta, is.numeric)
hiv_ltd_numeric <- hiv_ltd_delta[, numeric_columns]

# Compute the correlation matrix
correlation_matrix <- cor(hiv_ltd_numeric, use = "pairwise.complete.obs")

# Set the threshold for filtering
threshold <- 0.5

# Create a logical matrix for correlations above the threshold
high_correlations <- abs(correlation_matrix) > threshold

# Filter the correlation matrix
filtered_correlations <- correlation_matrix[high_correlations]

# Convert to a data frame for easier readability
filtered_correlations_df <- as.data.frame(as.table(filtered_correlations))

# Remove NA values and self-correlations
filtered_correlations_df <- na.omit(filtered_correlations_df)
filtered_correlations_df <- filtered_correlations_df[filtered_correlations_df$Var1 != filtered_correlations_df$Var2, ]

# Print the filtered correlations
print(filtered_correlations_df)




```



```{r, include=FALSE}

#simple linear regression models - getting residuals to check assumption of equal variance 

# Linear model for T-Cell Count
lm_LEU3N <- lm(LEU3N_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)
summary(lm_LEU3N)
confint(lm_LEU3N, level = 0.95)

# SLR for log-transformed Viral Load
lm_vload <- lm(log(VLOAD_delta + 1) ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)
summary(lm_vload)
confint(lm_vload, level = 0.95)  # Use the correct model for confidence intervals

# SLR for aggregated physical health
lm_phys <- lm(AGG_PHYS_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)
summary(lm_phys)
confint(lm_phys)

# SLR for aggregated mental health
lm_mental <- lm(AGG_MENT_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)
summary(lm_mental)
confint(lm_mental)
```

```{r,include=FALSE}

#calculate residuals and fitted values, of my models 

LEU3N_fitted<- predict(lm_LEU3N)
LEU3N_resid <-  resid(lm_LEU3N)

vload_fitted <- predict(lm_vload)
vload_resid <- resid(lm_vload)

phys_fitted <- predict(lm_phys)
phys_resid <- resid(lm_phys)


ment_fitted <- predict(lm_mental)
ment_resid  <- fitted(lm_mental)

 


```
Checking Assumptions of MANOVA

```{r,echo=FALSE}

# Plot Residuals vs Fitted LEU3N  (Homoscedasticity)
ggplot(data = NULL, aes(x = LEU3N_fitted, y = LEU3N_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Figure 1: CD4 T-cells Residuals vs Fitted", 
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()



#plot residuals vs fitted for Viral Load
ggplot(data = NULL, aes(x = vload_fitted, y = vload_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Figure 2: Viral load Residuals vs Fitted", 
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

#plot residuals for Physical 
ggplot(data = NULL, aes(x = phys_fitted, y = phys_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Figure 3: Physical Health Measurement Residuals vs Fitted", 
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()


ggplot(data = NULL, aes(x = ment_fitted, y = ment_resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Figure 3: PMental Health Measurement Residuals vs Fitted", 
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

```
Check Normality
```{r,echo=FALSE}

# Histogram of original VLOAD_delta
ggplot(hiv_ltd_delta, aes(x = log(VLOAD_delta)))+
  geom_histogram(bins = 30, fill = "blue", color = "white") +
  labs(title = "Histogram of log-transformed Viral Load Data",
       x = "Viral Load",
       y = "Frequency") +
  theme_minimal()

#logtransformed

# Histogram of LEU3N
ggplot(hiv_ltd_delta, aes(x = LEU3N_delta)) +
  geom_histogram(bins = 30, fill = "blue", color = "white") +
  labs(title = "Histogram of T-cell Change Data Points",
       x = "T-Cell Count",
       y = "Frequency") +
  theme_minimal()

# Histogram of Agg_Phys
ggplot(hiv_ltd_delta, aes(x = AGG_PHYS_delta)) +
  geom_histogram(bins = 30, fill = "blue", color = "white") +
  labs(title = "Histogram of Aggregated Physical Health Distribution",
       x = "Physical Health Change",
       y = "Frequency") +
  theme_minimal()

#Histogram of Agg_ment

ggplot(hiv_ltd_delta, aes(x = AGG_MENT_delta)) +
  geom_histogram(bins = 30, fill = "blue", color = "white") +
  labs(title = "Histogram of Aggregated Mental Health Distribution",
       x = "Mental Health",
       y = "Frequency") +
  theme_minimal()



```

```{r, echo=FALSE}
#qq Plots

# Q-Q Plot for T-Cell Count
qqnorm(lm_LEU3N$residuals, main = "Q-Q Plot of Residuals for T-Cell Count")
qqline(lm_LEU3N$residuals, col = "red")

# Q-Q Plot for log-transformed Viral Load
qqnorm(lm_vload$residuals, main = "Q-Q Plot of Residuals for Viral Load")
qqline(lm_vload$residuals, col = "red")

# Q-Q Plot for Aggregated Physical Health
qqnorm(lm_phys$residuals, main = "Q-Q Plot of Residuals for Aggregated Physical Health")
qqline(lm_phys$residuals, col = "red")

# Q-Q Plot for Aggregated Mental Health
qqnorm(lm_mental$residuals, main = "Q-Q Plot of Residuals for Aggregated Mental Health")
qqline(lm_mental$residuals, col = "red")



```

```{r, echo=FALSE}
#assumptions of MANOVA
# Check normality for each response variable
shapiro.test(resid(lm(LEU3N_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)))
shapiro.test(resid(lm(VLOAD_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)))
shapiro.test(resid(lm(AGG_PHYS_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)))
shapiro.test(resid(lm(AGG_MENT_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)))

# Q-Q Plots
par(mfrow = c(2, 2))
qqnorm(resid(lm(LEU3N_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)), main = "LEU3N_delta Q-Q Plot")
qqline(resid(lm(LEU3N_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)))

qqnorm(resid(lm(VLOAD_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)), main = "VLOAD_delta Q-Q Plot")
qqline(resid(lm(VLOAD_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)))

qqnorm(resid(lm(AGG_PHYS_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)), main = "AGG_PHYS_delta Q-Q Plot")
qqline(resid(lm(AGG_PHYS_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)))

qqnorm(resid(lm(AGG_MENT_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)), main = "AGG_MENT_delta Q-Q Plot")
qqline(resid(lm(AGG_MENT_delta ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)))



```
```{r,echo=FALSE}


# Fit the MANOVA model with additional response variables
manova_model_basic <- manova(cbind(LEU3N_delta, log(VLOAD_delta), AGG_PHYS_delta, AGG_MENT_delta) ~ drug_use_currently + drug_use_previously, data = hiv_ltd_delta)

# Summarize the MANOVA results using Pillai's trace, bonferroni's adjustment reuires p-value of 0.0125
summary(manova_model_basic, test = "Pillai")


```

```{r,echo=FALSE}

# Get the summary of the MANOVA model using Pillai's test
manova_summary_basic <- summary(manova_model_basic, test = "Pillai")

# Extract relevant results into a tidy format
manova_results_basic <- data.frame(
     Effect = rownames(manova_summary_basic$stats),
     Df = manova_summary_basic$stats[, "Df"],
     Pillai = manova_summary_basic$stats[, "Pillai"],
     approx_F = manova_summary_basic$stats[, "approx F"],
     num_Df = manova_summary_basic$stats[, "num Df"],
     den_Df = manova_summary_basic$stats[, "den Df"],
     p.value = manova_summary_basic$stats[, "Pr(>F)"]
)

# Remove the "Residuals" row for clarity
manova_results_basic <- manova_results_basic[-which(manova_results_basic$Effect == "Residuals"), ]

# Create a mapping of group names to labels
# Adjust the group labels according to your study
group_labels_basic <- c(
    drug_use_currently = "Current Drug Use",
    drug_use_previously = "Previous Drug Use"
)

# Replace Effect names with labels
manova_results_basic$Effect <- recode(manova_results_basic$Effect, !!!group_labels_basic)

#create table

manova_results_basic %>%
    kbl(caption = "MANOVA Results (Pillai's Trace) - Basic Model") %>%
    kable_styling(full_width = F)




```


Check assumptions of MANOVA on model looking at ADh levels

```{r}

# Fit a linear model with all predictors
# Fit SLR models for each outcome variable

# Full model for LEU3N_delta
lm_full_leu3n <- lm(LEU3N_delta ~ drug_use_currently + drug_use_previously + 
                     EDUCBAS_2 + EDUCBAS_3 + EDUCBAS_4 + 
                     EDUCBAS_5 + EDUCBAS_6 + EDUCBAS_7 + 
                     ADH_2_2 + ADH_2_3 + ADH_2_4, 
                     data = hiv_ltd_delta)

# Full model for log-transformed VLOAD_delta
lm_full_vload <- lm(log(VLOAD_delta + 1) ~ drug_use_currently + drug_use_previously + 
                     EDUCBAS_2 + EDUCBAS_3 + EDUCBAS_4 + 
                     EDUCBAS_5 + EDUCBAS_6 + EDUCBAS_7 + 
                     ADH_2_2 + ADH_2_3 + ADH_2_4, 
                     data = hiv_ltd_delta)

# Full model for AGG_PHYS_delta
lm_full_phys <- lm(AGG_PHYS_delta ~ drug_use_currently + drug_use_previously + 
                    EDUCBAS_2 + EDUCBAS_3 + EDUCBAS_4 + 
                    EDUCBAS_5 + EDUCBAS_6 + EDUCBAS_7 + 
                    ADH_2_2 + ADH_2_3 + ADH_2_4, 
                    data = hiv_ltd_delta)

# Full model for AGG_MENT_delta
lm_full_mental <- lm(AGG_MENT_delta ~ drug_use_currently + drug_use_previously + 
                      EDUCBAS_2 + EDUCBAS_3 + EDUCBAS_4 + 
                      EDUCBAS_5 + EDUCBAS_6 + EDUCBAS_7 + 
                      ADH_2_2 + ADH_2_3 + ADH_2_4, 
                      data = hiv_ltd_delta)







```

```{r}
#plot data to visualize normality and equal variance

# Function to plot residuals
plot_residuals <- function(model, title) {
  fitted_vals <- fitted(model)
  residuals <- resid(model)
  
  ggplot(data = NULL, aes(x = fitted_vals, y = residuals)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(title = title, x = "Fitted Values", y = "Residuals") +
    theme_minimal()
}

# Create plots
plot_residuals(lm_full_leu3n, "LEU3N_delta Residuals vs Fitted")
plot_residuals(lm_full_vload, "VLOAD_delta Residuals vs Fitted")
plot_residuals(lm_full_phys, "AGG_PHYS_delta Residuals vs Fitted")
plot_residuals(lm_full_mental, "AGG_MENT_delta Residuals vs Fitted")





```

```{r}

#shapiro wilke test for normality:

# Shapiro-Wilk test for normality of residuals
shapiro_tests <- list(
  shapiro.test(resid(lm_full_leu3n)),
  shapiro.test(resid(lm_full_vload)),
  shapiro.test(resid(lm_full_phys)),
  shapiro.test(resid(lm_full_mental))
)

# Print results
lapply(shapiro_tests, print)




```



```{r,echo=FALSE}

# Updated model with interaction terms for education
manova_model <- manova(cbind(LEU3N_delta, log(VLOAD_delta), AGG_PHYS_delta, AGG_MENT_delta) ~ 
                        drug_use_currently + 
                        drug_use_previously + 
                        EDUCBAS_2 + 
                        EDUCBAS_3 + 
                        EDUCBAS_4 + 
                        EDUCBAS_5 + 
                        EDUCBAS_6 + 
                        EDUCBAS_7 + 
                        ADH_2_2 + 
                        ADH_2_3 + 
                        ADH_2_4 + 
                        drug_use_currently * ADH_2_2 + 
                        drug_use_currently * ADH_2_3 + 
                        drug_use_currently * ADH_2_4 + 
                        drug_use_previously * ADH_2_2 + 
                        drug_use_previously * ADH_2_3 + 
                        drug_use_previously * ADH_2_4 +
                        drug_use_currently * EDUCBAS_2 + 
                        drug_use_currently * EDUCBAS_3 + 
                        drug_use_currently * EDUCBAS_4 + 
                        drug_use_currently * EDUCBAS_5 + 
                        drug_use_currently * EDUCBAS_6 + 
                        drug_use_currently * EDUCBAS_7 + 
                        drug_use_previously * EDUCBAS_2 + 
                        drug_use_previously * EDUCBAS_3 + 
                        drug_use_previously * EDUCBAS_4 + 
                        drug_use_previously * EDUCBAS_5 + 
                        drug_use_previously * EDUCBAS_6 + 
                        drug_use_previously * EDUCBAS_7,
                        data = hiv_ltd_delta)

# Summarize the MANOVA results using Pillai's trace
summary(manova_model, test = "Pillai")




```




```{r,echo=FALSE}



# Get the summary of the MANOVA model using Pillai's test
manova_summary <- summary(manova_model, test = "Pillai")

# Extract relevant results into a tidy format
manova_results <- data.frame(
     Effect = rownames(manova_summary$stats),
     Df = manova_summary$stats[, "Df"],
     Pillai = manova_summary$stats[, "Pillai"],
     approx_F = manova_summary$stats[, "approx F"],
     num_Df = manova_summary$stats[, "num Df"],
     den_Df = manova_summary$stats[, "den Df"],
     p.value = manova_summary$stats[, "Pr(>F)"]
)

# Remove the "Residuals" row for clarity
manova_results <- manova_results[-which(manova_results$Effect == "Residuals"), ]

# Create a nice table with kable


manova_results %>%
    kbl(caption = "MANOVA Results (Pillai's Trace)") %>%
    kable_styling(full_width = F)


```










```{r,echo=FALSE}
#scratch code not used in analysis- exploring data
#DQ check on VLOAD variable. logtransform may not be working due to missingness 

# Summary statistics for VLOAD_delta
summary(hiv_ltd_delta$VLOAD_delta)

# Check for missing values
sum(is.na(hiv_ltd_delta$VLOAD_delta))

# Check for negative values
negative_values <- hiv_ltd_delta$VLOAD_delta[hiv_ltd_delta$VLOAD_delta < 0]
print(negative_values)

# Count of negative values
num_negative_values <- length(negative_values)
cat("Number of negative values:", num_negative_values, "\n")






```












